---
title: "K-DBSCAN"
author: Philip Loewen, Arusan Surendiran
format: pdf
---



# DBSCAN Algorithm

This section defines the manual implementation of the DBSCAN algorithm. It utilizes a queue in data structure and a pairwise distance matrix to handle cluster expansion efficiently in R.

https://www.geeksforgeeks.org/dsa/basic-operations-for-queue-in-data-structure/


```{r}
manual_dbscan <- function(data, eps, minPts) {
  n <- nrow(data)
  cluster_assignment <- rep(0, n)  # 0 = unvisited, -1 = noise
  cluster_id <- 0
  
  # Compute pairwise Euclidean distances to define epsilon-neighborhoods 
  dist_mat <- as.matrix(dist(data, method = "euclidean"))
  
  for (i in 1:n) {
    # Skip point if it has already been visited
    if (cluster_assignment[i] != 0) next
    
    # Retrieve the epsilon-neighborhood for point i
    neighbors <- which(dist_mat[i, ] <= eps)
    
    # Check core point condition
    if (length(neighbors) < minPts) {
      cluster_assignment[i] <- -1 # Temporarily classify as a noise point
    } else {
      # Core point verified: create a new density-based cluster
      cluster_id <- cluster_id + 1
      cluster_assignment[i] <- cluster_id
      
      # Identify directly density-reachable points to start the chain
      seeds <- setdiff(neighbors, i)
      
      # Initialize a queue to establish density-connected points
      queue <- integer(n)
      queue_len <- length(seeds)
      
      if (queue_len > 0) {
        queue[1:queue_len] <- seeds
      }
      
      head <- 1
      tail <- queue_len
      
      # Expand the cluster to satisfy the Maximality requirement
      while (head <= tail) {
        current_p <- queue[head]
        head <- head + 1 
        
        # If previously classified as noise, it is a border point
        if (cluster_assignment[current_p] == -1) {
          cluster_assignment[current_p] <- cluster_id
        }
        
        # If unvisited, process for density-reachability
        if (cluster_assignment[current_p] == 0) {
          cluster_assignment[current_p] <- cluster_id
          
          # Retrieve the epsilon-neighborhood of the current point
          new_neighbors <- which(dist_mat[current_p, ] <= eps)
          
          # If this point is also a core point, expand the reachable chain
          if (length(new_neighbors) >= minPts) {
            # Extract only unvisited or noise points
            valid_new <- new_neighbors[cluster_assignment[new_neighbors] <= 0]
            num_new <- length(valid_new)
            
            if (num_new > 0) {
              queue[(tail + 1):(tail + num_new)] <- valid_new
              tail <- tail + num_new
            }
          }
        }
      }
    }
  }
  return(cluster_assignment)
}
```

# Datasets

This section handles the retrieval and preprocessing of the real-world datasets from the UCI Machine Learning Repository.

```{r}
library(remotes)
#remotes::install_github("coatless-rpkg/ucimlrepo")
library(ucimlrepo)
library(here)

# Abalone Dataset
abalone_uci <- fetch_ucirepo(name = "abalone")
abalone <- abalone_uci$data$original

# Prepare and scale Abalone for clustering
abalone_numeric <- abalone[, 2:8]
abalone_features <- scale(abalone_numeric)

# MAGIC Gamma Telescope Dataset
MAGIC_uci <- fetch_ucirepo(id=159)
magic <- MAGIC_uci$data$original

# Prepare and scale Abalone for clustering
magic_unscaled <- magic[, 1:10]
magic_features <- scale(magic_unscaled)

# Shuttle Data

shuttle_train <- read.table("data/shuttle.trn", header = FALSE, sep = " ")
shuttle_test <- read.table("data/shuttle.tst", header = FALSE, sep = " ")
statlog_shuttle <- rbind(shuttle_train, shuttle_test)

# Separate features from labels, the last column (V10) is the class label
shuttle_unscaled <- statlog_shuttle[, 1:9]
shuttle_features <- scale(shuttle_unscaled)

# MoCap Hand postures Data
mocap <- read.csv("data/mocap-hand-postures.csv")
mocap_unscaled <- mocap[, 3:11]
mocap_features <- scale(mocap_unscaled)
```



# Simulation

The simulation evaluates the algorithm's performance on non-linear geometries (sine curves, rings, and arcs) and its ability to distinguish clusters from background noise. We demonstrate the advantages of DBSCAN clustering in comparison to k-means clustering.

```{r}
library(dbscan)
library(ggplot2)
library(MASS)
set.seed(2201)

# Define the cluster centers and sizes
centers <- matrix(c(25, 25, -25, -25, 25, -20, -20, 20), ncol = 2, byrow = TRUE) 
sizes <- c(1000, 1000, 1000, 1000, 1000, 1000, 1000, 1000, 1000, 1000, 250, 100)

# Generate synthetic clusters with different characteristics
cluster1 <- mvrnorm(n = sizes[1], mu = centers[1,], Sigma = matrix(c(9, 2, 2, 9), ncol = 2))
cluster2 <- mvrnorm(n = sizes[2], mu = centers[2,], Sigma = matrix(c(12, 0.5, 0.5, 4), ncol = 2))
cluster3 <- mvrnorm(n = sizes[3], mu = centers[3,], Sigma = matrix(c(6, 0.3, 0.3, 18), ncol = 2))
cluster4 <- mvrnorm(n = sizes[4], mu = centers[4,], Sigma = matrix(c(15, 0.1, 0.1, 15), ncol = 2))

# Cluster 5: Sine curve distribution
x_values <- seq(-40, -20, length.out = 1000)
y_values <- sin(x_values) * 2 + rnorm(1000, mean = 0, sd = 0.2) # Adding some noise to the sine curve
cluster5 <- cbind(x_values, y_values)

# Cluster 6: Cluster
cluster6 <- mvrnorm(n = 1000, mu = c(-3, -7), Sigma = matrix(c(7, 1.3, 1.3, 5), ncol = 2))

# Cluster 7: Points along an arc
theta <- runif(1000, min = pi, max = 2 * pi)
r <- 5
cluster7 <- cbind(r * cos(theta) + 5, r * sin(theta) + 15)

# Cluster 8: Points in a "moon" shape
theta <- runif(1000, min = 0, max = pi)
r <- 10 + runif(1000, min = -1, max = 1)
cluster8 <- cbind(r * cos(theta) + 30, r * sin(theta))

# Cluster 9: Points in a "ring" shape around all other clusters
theta <- runif(1000, min = 0, max = 2 * pi)
r_inner <- 48
r_outer <- 52
r <- sqrt(runif(1000, min = r_inner^2, max = r_outer^2))
cluster9 <- cbind(r * cos(theta), r * sin(theta))

# Noise and Outliers
noise <- cbind(runif(250, -60, 60), runif(250, -60, 60))
outliers <- cbind(rnorm(100, 0, 20), rnorm(100, 0, 20))

data_full <- rbind(cluster1, cluster2, cluster3, cluster4, cluster5, 
                   cluster6, cluster7, cluster8, cluster9, noise, outliers)
df_full <- as.data.frame(data_full)
colnames(df_full) <- c("X", "Y")
```


## Execution and Comparison

```{r}
set.seed(2201)
# Run Manual DBSCAN
results_dbscan <- manual_dbscan(df_full, eps = 2.5, minPts = 4)
df_full$Cluster <- as.factor(results_dbscan)

# Run K-means for Comparison
km_result <- kmeans(df_full, centers = 10, nstart = 25)
df_full$Kmeans_Cluster <- as.factor(km_result$cluster)
```



## Visualization

```{r}

# Extract unique clusters excluding the noise (-1)
db_clusters <- unique(results_dbscan[results_dbscan != -1])
db_colors <- setNames(scales::hue_pal()(length(db_clusters)), db_clusters)
db_palette <- c("-1" = "grey70", db_colors)

  
# DBSCAN Results
ggplot(df_full, aes(x = X, y = Y, color = Cluster)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = db_palette) +
  theme_minimal() +
  labs(title = "Manual DBSCAN", subtitle = "Grey = Noise (-1)") +
  theme(legend.position = "none")

# K-means Results
ggplot(df_full, aes(x = X, y = Y, color = Kmeans_Cluster)) +
  geom_point(alpha = 0.5, size = 0.8) +
  theme_minimal() +
  labs(title = "K-means Comparison", subtitle = "Note: Noise forced into clusters") +
  theme(legend.position = "none")
```





# Application of Algorithm


```{r}
#| include: false
#| eval: false
manual_dbscan(data, eps = 0.05, minPts = 4)
```


```{r}
#| include: false
#| eval: false

colnames(abalone) <- c("Sex", "Length", "Diameter", "Height", "Whole_w", 
                       "Shucked_w", "Viscera_w", "Shell_w", "Rings")

# Step 1: Remove 'Sex' for clustering or use model.matrix to dummy code it
# For a manual implementation, let's keep it simple and just use the numeric columns (2:8)
abalone_numeric <- abalone[, 2:8]

# Step 2: SCALE THE DATA
# This is non-negotiable for DBSCAN!
abalone_scaled <- scale(abalone_numeric)
results <- manual_dbscan(abalone_scaled, eps = 1, minPts = 4)

print(table(results))
```


# K-means++ Algorithm

K-means++ improves upon the standard k-means algorithm by providing a specific method for choosing initial cluster centers. Standard k-means often relies on random initialization, which can lead to poor local optima. The k-means++ approach spreads out the initial centers before starting the iterative assignment and update steps.

## Implementation

The following implementation includes the probabilistic seeding process and the iterative refinement loop.

```{r}
# Function to compute Euclidean distance between a point and a set of centers
get_min_dist_sq <- function(data, centers) {
  # Calculate squared distances from each point to the nearest existing center
  apply(data, 1, function(point) {
    min(colSums((t(centers) - point)^2))
  })
}

manual_kmeans_pp <- function(data, k, max_iter = 100) {
  data_mat <- as.matrix(data)
  n <- nrow(data_mat)
  
  # --- Step 1: K-means++ Seeding ---
  
  # Choose the first center uniformly at random
  centers <- matrix(NA, nrow = k, ncol = ncol(data_mat))
  centers[1, ] <- data_mat[sample(1:n, 1), ]
  
  if (k > 1) {
    for (i in 2:k) {
      # Compute squared distances to the nearest center already chosen
      dist_sq <- get_min_dist_sq(data_mat, centers[1:(i-1), , drop = FALSE])
      
      # Select next center with probability proportional to D(x)^2
      # This ensures centers are spread out across the feature space
      probs <- dist_sq / sum(dist_sq)
      next_idx <- sample(1:n, 1, prob = probs)
      centers[i, ] <- data_mat[next_idx, ]
    }
  }
  
  # --- Step 2: Standard K-means Iteration ---
  
  assignments <- rep(0, n)
  prev_assignments <- rep(-1, n)
  iter <- 0
  
  while (!all(assignments == prev_assignments) && iter < max_iter) {
    prev_assignments <- assignments
    iter <- iter + 1
    
    # Assignment Step: Assign each point to the nearest center
    dist_to_centers <- as.matrix(dist(rbind(centers, data_mat)))
    # Extract distances between points and centers only
    dist_matrix <- dist_to_centers[(k + 1):(n + k), 1:k]
    assignments <- apply(dist_matrix, 1, which.min)
    
    # Update Step: Recompute centers as the mean of assigned points
    for (j in 1:k) {
      if (sum(assignments == j) > 0) {
        centers[j, ] <- colMeans(data_mat[assignments == j, , drop = FALSE])
      }
    }
  }
  
  return(list(cluster = assignments, centers = centers, iterations = iter))
}
```

## Simulation and Evaluation

We reuse the previous synthetic dataset to evaluate how effectively k-means++ identifies the underlying structure.

```{r}
#| cache: true
set.seed(2201)
# Run Manual K-means++
# We choose k=10 to match the number of intended clusters in the generation
k_clusters <- 10
results_kmpp <- manual_kmeans_pp(df_full[, c("X", "Y")], k = k_clusters)
df_full$Kmpp_Cluster <- as.factor(results_kmpp$cluster)

centers_df <- as.data.frame(results_kmpp$centers)
colnames(centers_df) <- c("X", "Y") 
```

## Visualization

The plot below displays the final cluster assignments. The larger black points represent the final centroids calculated by the algorithm.

```{r}
# Kmpp-means Results
ggplot(df_full, aes(x = X, y = Y, color = Kmpp_Cluster)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_point(data = centers_df, aes(x = X, y = Y), 
             color = "black", size = 3) +
  theme_minimal() +
  labs(title = "K-means++ Comparison", subtitle = "Note: Noise forced into clusters") +
  theme(legend.position = "none")
```

## Application to Abalone dataset

This section applies the k-means++ implementation to the numeric features of the Abalone dataset.

```{r}
# Assuming the 'abalone_scaled' object exists from previous preprocessing
# If not, ensure the data is loaded and scaled as shown in the DBSCAN section.
if (exists("abalone_scaled")) {
  abalone_results <- manual_kmeans_pp(abalone_scaled, k = 3)
  table(abalone_results$cluster)
} else {
  print("Scaled Abalone data not found. Please run the preprocessing steps.")
}
```

# K-DBSCAN Algorithm

The primary goal of the K-DBSCAN hybrid approach is to process data in localized segments and then unify density-connected clusters that cross the boundaries of those segments. By partitioning the space with K-means, the algorithm only performs exhaustive distance checks between points within the same group or between adjacent groups. 

## Implementation

The following implementation integrates the `manual_dbscan` and `manual_kmeans_pp` methods.

```{r}
single_link_dist <- function(clu_1, clu_2) {
  clu_1 <- as.matrix(clu_1)
  clu_2 <- as.matrix(clu_2)

  if (nrow(clu_1) == 0 || nrow(clu_2) == 0) {
    return(Inf)
  }

  dist_mat <- as.matrix(dist(rbind(clu_1, clu_2)))
  n1 <- nrow(clu_1)
  n2 <- nrow(clu_2)
  return(min(dist_mat[1:n1, (n1 + 1):(n1 + n2), drop = FALSE]))
}

cluster_dist <- function(clu_1, clu_2) {
  clu_1 <- as.matrix(clu_1)
  clu_2 <- as.matrix(clu_2)

  if (nrow(clu_1) == 0 || nrow(clu_2) == 0) {
    return(Inf)
  }

  center_1 <- colMeans(clu_1)
  center_2 <- colMeans(clu_2)

  dis_KCij <- sqrt(sum((center_1 - center_2)^2))
  dis_ri <- max(sqrt(rowSums((clu_1 - matrix(center_1, nrow = nrow(clu_1), ncol = ncol(clu_1), byrow = TRUE))^2)))
  dis_rj <- max(sqrt(rowSums((clu_2 - matrix(center_2, nrow = nrow(clu_2), ncol = ncol(clu_2), byrow = TRUE))^2)))

  return(dis_KCij - dis_ri - dis_rj)
}

k_dbscan <- function(D, k, max_iter, eps, minPts) {
  # Step 1: Normalization of D
  D_mat <- as.matrix(scale(D))
  n <- nrow(D_mat)

  if (n == 0) {
    return(integer(0))
  }

  # Step 2: Data grouping with K-means Algorithm
  km_results <- manual_kmeans_pp(D_mat, k = k, max_iter = max_iter)
  group_id <- km_results$cluster

  # Step 3: Run DBSCAN inside each K-means group
  group_points <- vector("list", k)
  numclu <- integer(k)

  for (i in seq_len(k)) {
    group_idx <- which(group_id == i)
    if (length(group_idx) == 0) {
      group_points[[i]] <- list()
      next
    }

    group_data <- D_mat[group_idx, , drop = FALSE]
    local_labels <- manual_dbscan(group_data, eps, minPts)
    local_cluster_ids <- sort(unique(local_labels[local_labels > 0]))

    numclu[i] <- length(local_cluster_ids)
    group_points[[i]] <- lapply(local_cluster_ids, function(local_cluster) {
      group_idx[local_labels == local_cluster]
    })
  }

  # Step 4: Merge border clusters
  flag <- vector("list", k)
  for (i in seq_len(k)) {
    flag[[i]] <- rep(0L, numclu[i])
  }

  counter <- 1L

  for (a in seq_len(k - 1L)) {
    for (b in seq_len(k)) {
      if (numclu[a] == 0L || numclu[b] == 0L) next

      dis_k <- cluster_dist(D_mat[group_id == a, , drop = FALSE],
                            D_mat[group_id == b, , drop = FALSE])
      if (dis_k > eps) next

      for (c in seq_len(numclu[a])) {
        for (d in seq_len(numclu[b])) {
          dis_db <- single_link_dist(
            D_mat[group_points[[a]][[c]], , drop = FALSE],
            D_mat[group_points[[b]][[d]], , drop = FALSE]
          )

          if (dis_db <= eps) {
            if (flag[[a]][c] == 0L && flag[[b]][d] == 0L) {
              flag[[a]][c] <- counter
              flag[[b]][d] <- counter
              counter <- counter + 1L
            } else {
              new_flag <- max(flag[[a]][c], flag[[b]][d])
              flag[[a]][c] <- new_flag
              flag[[b]][d] <- new_flag
            }
          }
        }
      }
    }
  }

  for (i in seq_len(k)) {
    if (numclu[i] == 0L) next
    for (c in seq_len(numclu[i])) {
      if (flag[[i]][c] == 0L) {
        flag[[i]][c] <- counter
        counter <- counter + 1L
      }
    }
  }

  # Step 5: Run DBSCAN again inside each flagged component
  final_labels <- rep(-1L, n)
  final_cluster_id <- 0L

  for (i in seq_len(counter - 1L)) {
    component_points <- integer(0)

    for (g in seq_len(k)) {
      if (numclu[g] == 0L) next
      for (c in seq_len(numclu[g])) {
        if (flag[[g]][c] == i) {
          component_points <- c(component_points, group_points[[g]][[c]])
        }
      }
    }

    component_points <- sort(unique(component_points))
    if (length(component_points) == 0) next

    flagged_data <- D_mat[component_points, , drop = FALSE]
    flagged_result <- manual_dbscan(flagged_data, eps, minPts)

    for (local_cluster in sort(unique(flagged_result[flagged_result > 0]))) {
      final_cluster_id <- final_cluster_id + 1L
      final_labels[component_points[flagged_result == local_cluster]] <- final_cluster_id
    }
  }

  return(final_labels)
}
```


```{r}
results_kdbscan <- k_dbscan(df_full[, c("X", "Y")], 10, 4, 0.12, 5)
df_full$kdbs_cluster <- factor(results_kdbscan)

real_clusters <- sort(unique(results_kdbscan[results_kdbscan > 0]))

ggplot(df_full, aes(x = X, y = Y, color = kdbs_cluster)) +
  geom_point(size = 0.5, alpha = 0.7) +
  scale_color_manual(
    values = c("-1" = "grey70",
               setNames(scales::hue_pal()(length(real_clusters)),
                        as.character(real_clusters)))
  ) +
  labs(title = "K-DBSCAN clustering", subtitle = "Grey = Noise (-1)") +
  theme_minimal() +
  theme(legend.position = "none")
```


## Datasets Implementation


```{r}
# 1. Define Datasets and Parameters
datasets <- list(
  Abalone = abalone_features,
  MAGIC   = magic_features,
  Shuttle = shuttle_features,
  MoCap   = mocap_features
)

params <- list(
  Abalone = list(eps = 0.035, minPts = 4),
  MAGIC   = list(eps = 0.06,  minPts = 4),
  Shuttle = list(eps = 0.02,  minPts = 5),
  MoCap   = list(eps = 0.007,  minPts = 4)
)
```


```{r}
# --- 1. Helper Functions ---
find_uf <- function(parent, x) {
  while (parent[x] != x) x <- parent[x]
  return(x)
}

union_uf <- function(parent, x, y) {
  rx <- find_uf(parent, x)
  ry <- find_uf(parent, y)
  if (rx != ry) parent[rx] <- ry
  return(parent)
}

# Safely calculate the max distance from a center (cluster radius)
get_radius <- function(data_mat, center_vec) {
  if (nrow(data_mat) == 0) return(0)
  diffs <- sweep(data_mat, 2, center_vec, "-")
  return(max(sqrt(rowSums(diffs^2))))
}

# Fast cross-distance using linear algebra to avoid R loops
fast_min_dist <- function(A, B) {
  A <- as.matrix(A)
  B <- as.matrix(B)
  if (nrow(A) == 0 || nrow(B) == 0) return(Inf)
  
  D2 <- outer(rowSums(A^2), rowSums(B^2), "+") - 2 * tcrossprod(A, B)
  D2[D2 < 0] <- 0 
  return(sqrt(min(D2)))
}

# --- 2. Memory-Safe Manual DBSCAN (The O(N^2) Baseline) ---
manual_dbscan_safe <- function(D, eps, minPts) {
  D <- as.matrix(D)
  n <- nrow(D)
  if (n == 0) return(integer(0))
  
  labels <- rep(0L, n)
  cluster_id <- 0L
  
  for (i in seq_len(n)) {
    if (labels[i] != 0L) next
    
    diffs <- sweep(D, 2, D[i, ], "-")
    dists <- sqrt(rowSums(diffs^2))
    neighbors <- which(dists <= eps)
    
    if (length(neighbors) < minPts) {
      labels[i] <- -1L
    } else {
      cluster_id <- cluster_id + 1L
      labels[i] <- cluster_id
      seeds <- setdiff(neighbors, i)
      
      while (length(seeds) > 0) {
        cp <- seeds[1]
        seeds <- seeds[-1]
        
        if (labels[cp] == -1L) labels[cp] <- cluster_id
        if (labels[cp] != 0L) next
        
        labels[cp] <- cluster_id
        diffs_seed <- sweep(D, 2, D[cp, ], "-")
        dists_seed <- sqrt(rowSums(diffs_seed^2))
        seed_neighbors <- which(dists_seed <= eps)
        
        if (length(seed_neighbors) >= minPts) {
          new_seeds <- setdiff(seed_neighbors, c(seeds, i, cp))
          if (length(new_seeds) > 0) seeds <- c(seeds, new_seeds)
        }
      }
    }
  }
  return(labels)
}

# --- 3. Optimized K-DBSCAN ---
k_dbscan_opt <- function(D, eps, minPts) {
  D_mat <- as.matrix(scale(D))
  n <- nrow(D_mat)
  if (n < 2) return(rep(-1L, n))
  
  k <- max(1, min(floor(sqrt(n)), n - 1))
  
  km <- suppressWarnings(kmeans(D_mat, centers = k, iter.max = 4))
  group_id <- km$cluster
  centers <- km$centers
  
  group_points <- vector("list", k)
  numclu <- integer(k)
  group_radii <- numeric(k)
  sub_centers <- vector("list", k)
  sub_radii <- vector("list", k)
  
  for (i in seq_len(k)) {
    idx <- which(group_id == i)
    if (length(idx) == 0) next
    
    group_data <- D_mat[idx, , drop = FALSE]
    group_radii[i] <- get_radius(group_data, centers[i, ])
    
    local_labels <- manual_dbscan_safe(group_data, eps, minPts)
    u_labels <- sort(unique(local_labels[local_labels > 0]))
    numclu[i] <- length(u_labels)
    
    group_points[[i]] <- lapply(u_labels, function(l) idx[local_labels == l])
    sub_centers[[i]] <- lapply(group_points[[i]], function(p) colMeans(D_mat[p, , drop = FALSE]))
    sub_radii[[i]] <- mapply(function(p, ctr) get_radius(D_mat[p, , drop = FALSE], ctr), 
                             group_points[[i]], sub_centers[[i]])
  }

  total_subclu <- sum(numclu)
  if (total_subclu == 0) return(rep(-1L, n))
  parent <- seq_len(total_subclu)
  
  sub_id_map <- vector("list", k)
  curr <- 0
  for (g in seq_len(k)) { 
    if(numclu[g] > 0) { 
      sub_id_map[[g]] <- (curr + 1):(curr + numclu[g])
      curr <- curr + numclu[g] 
    } 
  }
  
  g_dist <- as.matrix(dist(centers))
  g_adj <- (g_dist - outer(group_radii, group_radii, "+")) <= eps
  g_adj[lower.tri(g_adj, diag = TRUE)] <- FALSE
  pairs <- which(g_adj, arr.ind = TRUE)
  
  if (nrow(pairs) > 0) {
    for (idx in seq_len(nrow(pairs))) {
      g1 <- pairs[idx, 1]
      g2 <- pairs[idx, 2]
      
      for (c1 in seq_len(numclu[g1])) {
        for (c2 in seq_len(numclu[g2])) {
          d_dc <- sqrt(sum((sub_centers[[g1]][[c1]] - sub_centers[[g2]][[c2]])^2))
          if ((d_dc - (sub_radii[[g1]][c1] + sub_radii[[g2]][c2])) <= eps) {
            
            if (fast_min_dist(D_mat[group_points[[g1]][[c1]], ], D_mat[group_points[[g2]][[c2]], ]) <= eps) {
              parent <- union_uf(parent, sub_id_map[[g1]][c1], sub_id_map[[g2]][c2])
            }
            
          }
        }
      }
    }
  }

  roots <- unique(vapply(seq_len(total_subclu), find_uf, integer(1), parent = parent))
  final_labels <- rep(-1L, n)
  cid <- 0
  
  for (r in roots) {
    pts <- unlist(lapply(seq_len(k), function(g) {
      if(numclu[g] > 0) {
        matches <- which(vapply(sub_id_map[[g]], find_uf, 1, parent=parent) == r)
        if (length(matches) > 0) return(unlist(group_points[[g]][matches]))
      }
      return(NULL)
    }))
    
    if (length(pts) > 0) { 
      cid <- cid + 1
      final_labels[pts] <- cid 
    }
  }
  return(final_labels)
}
```


```{r}
# --- 1. Simulation Setup ---
test_sizes <- c(4000, 8000, 16000, 32000, 64000)
eps_sim <- 0.15
minPts_sim <- 5

benchmark_results <- data.frame(
  N = numeric(),
  DBSCAN_Time = numeric(),
  K_DBSCAN_Time = numeric()
)

# --- 2. The Benchmarking Loop ---
for (n in test_sizes) {
  cat("\nBenchmarking N =", n, "...\n")
  
  # Generate 4 distinct Gaussian blobs in 2D space
  set.seed(42)
  blob_centers <- matrix(c(-2,-2,  2,2,  -2,2,  2,-2), ncol = 2, byrow = TRUE)
  points_per_blob <- floor(n / 4)
  
  data_list <- lapply(1:4, function(i) {
    matrix(rnorm(points_per_blob * 2, mean = blob_centers[i,], sd = 0.4), ncol = 2)
  })
  
  sim_data <- do.call(rbind, data_list)
  # Standardize as expected by the algorithms
  sim_data <- scale(sim_data) 
  
  # Benchmark Standard DBSCAN
  cat("  Running Standard DBSCAN... ")
  t_dbscan <- system.time({
    res_db <- manual_dbscan_safe(sim_data, eps_sim, minPts_sim)
  })
  cat(round(t_dbscan["elapsed"], 3), "s\n")
  
  # Benchmark K-DBSCAN
  cat("  Running K-DBSCAN... ")
  t_kdbscan <- system.time({
    res_kdb <- k_dbscan_opt(sim_data, eps_sim, minPts_sim)
  })
  cat(round(t_kdbscan["elapsed"], 3), "s\n")
  
  # Store results
  benchmark_results <- rbind(benchmark_results, data.frame(
    N = n,
    DBSCAN_Time = t_dbscan["elapsed"],
    K_DBSCAN_Time = t_kdbscan["elapsed"]
  ))
}

# --- 3. Print Final Results ---
cat("\nFinal Benchmark Results:\n")
print(benchmark_results)
```

```{r}
#| fig-cap: Diagram for comparison of execution time in the proposed method, K-DBSCAN, and standard DBSCAN.


library(ggplot2)

# 1. Reshape the data into a long format for ggplot
df_lines <- data.frame(
  N = rep(benchmark_results$N, 2),
  Time = c(benchmark_results$DBSCAN_Time, benchmark_results$K_DBSCAN_Time),
  Algorithm = factor(
    rep(c("Standard DBSCAN", "K-DBSCAN"), each = nrow(benchmark_results)),
    levels = c("Standard DBSCAN", "K-DBSCAN")
  )
)

# 2. Create the plot
ggplot(df_lines, aes(x = N, y = Time, color = Algorithm, shape = Algorithm)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Standard DBSCAN" = "red", "K-DBSCAN" = "blue")) +
  scale_shape_manual(values = c("Standard DBSCAN" = 16, "K-DBSCAN" = 17)) +
  labs(
    title = "Comparing Standard DBSCAN vs K-DBSCAN for Large Datasets",
    x = "Number of Data Points (N)",
    y = "Execution Time (Seconds)"
  ) +
  scale_x_continuous(limits = c(min(df_lines$N), max(df_lines$N))) +
  scale_y_continuous(limits = c(0, max(df_lines$Time) * 1.1)) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = c(0.15, 0.85), # Positions legend in the top-left
    legend.title = element_blank(),
    legend.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "lightgray", linetype = "dotted")
  )
```



```{r}
library(cluster)     
library(clusterSim)  

evaluate_packaged_metrics <- function(data, labels, max_sample = 5000) {
  # 1. Filter out noise points (-1) and unassigned points (0)
  valid_idx <- labels > 0
  
  if (sum(valid_idx) == 0 || length(unique(labels[valid_idx])) < 2) {
    return(c(DB = NA, SI = NA))
  }
  
  data_c <- data[valid_idx, , drop = FALSE]
  labels_c <- labels[valid_idx]
  
  # 2. Davies-Bouldin Index (FIXED: parameter is 'centrotypes')
  db_res <- index.DB(data_c, as.integer(labels_c), centrotypes = "centroids")
  db_val <- db_res$DB
  
  # 3. Silhouette Index (Sampled for memory safety)
  if (nrow(data_c) > max_sample) {
    set.seed(42)
    sample_idx <- sample(seq_len(nrow(data_c)), max_sample)
    data_c <- data_c[sample_idx, ]
    labels_c <- labels_c[sample_idx]
  }
  
  dist_mat <- dist(data_c)
  sil_res <- silhouette(as.integer(labels_c), dist_mat)
  si_val <- mean(sil_res[, "sil_width"])
  
  return(c(DB = db_val, SI = si_val))
}
```


```{r}

## nO NOISE
test_sizes <- c(4000, 8000, 16000, 32000)
eps_sim <- 0.15
minPts_sim <- 5

metrics_results <- data.frame(
  N = numeric(),
  DBSCAN_DB = numeric(),
  DBSCAN_SI = numeric(),
  K_DBSCAN_DB = numeric(),
  K_DBSCAN_SI = numeric()
)

for (n in test_sizes) {
  cat("\nEvaluating Metrics for N =", n, "...\n")
  
  # Generate simulated clusters
  set.seed(42)
  blob_centers <- matrix(c(-2,-2,  2,2,  -2,2,  2,-2), ncol = 2, byrow = TRUE)
  points_per_blob <- floor(n / 4)
  
  data_list <- lapply(1:4, function(i) {
    matrix(rnorm(points_per_blob * 2, mean = blob_centers[i,], sd = 0.4), ncol = 2)
  })
  sim_data <- scale(do.call(rbind, data_list))
  
  # Get cluster labels
  res_db <- manual_dbscan_safe(sim_data, eps_sim, minPts_sim)
  res_kdb <- k_dbscan_opt(sim_data, eps_sim, minPts_sim)
  
  # Evaluate metrics
  metrics_db <- evaluate_packaged_metrics(sim_data, res_db)
  metrics_kdb <- evaluate_packaged_metrics(sim_data, res_kdb)
  
  # Store results
  metrics_results <- rbind(metrics_results, data.frame(
    N = n,
    DBSCAN_DB = metrics_db["DB"],
    DBSCAN_SI = metrics_db["SI"],
    K_DBSCAN_DB = metrics_kdb["DB"],
    K_DBSCAN_SI = metrics_kdb["SI"]
  ))
}

cat("\n========================================\n")
cat("FINAL METRICS COMPARISON:\n")
print(metrics_results)
```


```{r}
# --- 1. Define Sizes and Scaling Logic ---
test_sizes <- c(4000, 8000, 16000, 32000)
noise_ratio <- 0.10

# Helper to calculate the noise box limit based on N to keep density constant
# Reference: N=4000 has limit=4. Area scales linearly with N.
get_noise_limit <- function(n) {
  return(4 * sqrt(n / 4000))
}

# --- 2. Generate the Data List ---
datasets <- list()

for (n in test_sizes) {
  set.seed(n) # Unique seed per size
  
  n_noise <- floor(n * noise_ratio)
  n_signal <- n - n_noise
  pts_per_blob <- floor(n_signal / 4)
  
  # Signal: 4 Gaussian Blobs
  blob_centers <- matrix(c(-2,-2, 2,2, -2,2, 2,-2), ncol = 2, byrow = TRUE)
  signal_data <- do.call(rbind, lapply(1:4, function(i) {
    matrix(rnorm(pts_per_blob * 2, mean = blob_centers[i,], sd = 0.4), ncol = 2)
  }))
  
  # Noise: Adaptable Space (Expanding the box to keep density low)
  limit <- get_noise_limit(n)
  noise_data <- matrix(runif(n_noise * 2, min = -limit, max = limit), ncol = 2)
  
  # Combine, Shuffle, and Scale
  full_data <- rbind(signal_data, noise_data)
  full_data <- full_data[sample(nrow(full_data)), ]
  
  # Store in list with metadata
  datasets[[paste0("N", n)]] <- list(
    data = scale(full_data),
    limit = limit,
    n_noise = n_noise
  )
}
```


```{r}
# Final storage for results
final_metrics <- data.frame(
  N = numeric(),
  Noise_Limit = numeric(),
  DBSCAN_DB = numeric(),
  DBSCAN_SI = numeric(),
  K_DBSCAN_DB = numeric(),
  K_DBSCAN_SI = numeric()
)

# Shared Parameters
eps_val <- 0.15
minPts_val <- 10

for (name in names(datasets)) {
  # Extract data from our pre-generated list
  obj <- datasets[[name]]
  current_data <- obj$data
  n_val <- nrow(current_data)
  
  cat("\nProcessing Dataset:", name, "| Noise Box Limit:", round(obj$limit, 2), "\n")
  
  # Run clustering (Assuming manual_dbscan_safe and k_dbscan_opt are loaded)
  res_db <- manual_dbscan_safe(current_data, eps_val, minPts_val)
  res_kdb <- k_dbscan_opt(current_data, eps_val, minPts_val)
  
  # Evaluate Quality (Filtering out -1 noise automatically)
  m_db <- evaluate_packaged_metrics(current_data, res_db)
  m_kdb <- evaluate_packaged_metrics(current_data, res_kdb)
  
  # Append to results table
  final_metrics <- rbind(final_metrics, data.frame(
    N = n_val,
    Noise_Limit = obj$limit,
    DBSCAN_DB = m_db["DB"],
    DBSCAN_SI = m_db["SI"],
    K_DBSCAN_DB = m_kdb["DB"],
    K_DBSCAN_SI = m_kdb["SI"]
  ))
}

cat("\n========================================\n")
cat("CLEAN COMPARISON (CONSTANT NOISE DENSITY):\n")
print(final_metrics)
```



```{r}
test_sizes <- c(4000, 8000, 16000, 32000)
eps_sim <- 0.15
minPts_sim <- 5
noise_ratio <- 0.10 # 10% of the data will be random noise

metrics_results <- data.frame(
  N = numeric(),
  DBSCAN_DB = numeric(),
  DBSCAN_SI = numeric(),
  K_DBSCAN_DB = numeric(),
  K_DBSCAN_SI = numeric()
)

for (n in test_sizes) {
  cat("\nEvaluating Metrics for N =", n, "with 10% noise...\n")
  
  # 1. Calculate split between signal and noise
  n_noise <- floor(n * noise_ratio)
  n_signal <- n - n_noise
  points_per_blob <- floor(n_signal / 4)
  
  # 2. Generate the 4 signal blobs (Gaussian)
  set.seed(42)
  blob_centers <- matrix(c(-2,-2,  2,2,  -2,2,  2,-2), ncol = 2, byrow = TRUE)
  
  data_list <- lapply(1:4, function(i) {
    matrix(rnorm(points_per_blob * 2, mean = blob_centers[i,], sd = 0.4), ncol = 2)
  })
  signal_data <- do.call(rbind, data_list)
  
  # 3. Generate the background noise (Uniform distribution across the blob bounding box)
  noise_data <- matrix(runif(n_noise * 2, min = -4, max = 4), ncol = 2)
  
  # 4. Combine, shuffle, and scale
  sim_data <- rbind(signal_data, noise_data)
  sim_data <- sim_data[sample(nrow(sim_data)), ] # Shuffle rows randomly
  sim_data <- scale(sim_data)
  
  # Run the clustering algorithms
  res_db <- manual_dbscan_safe(sim_data, eps_sim, minPts_sim)
  res_kdb <- k_dbscan_opt(sim_data, eps_sim, minPts_sim)
  
  # Evaluate metrics (our function automatically filters out the -1 noise labels)
  metrics_db <- evaluate_packaged_metrics(sim_data, res_db)
  metrics_kdb <- evaluate_packaged_metrics(sim_data, res_kdb)
  
  # Store results
  metrics_results <- rbind(metrics_results, data.frame(
    N = n,
    DBSCAN_DB = metrics_db["DB"],
    DBSCAN_SI = metrics_db["SI"],
    K_DBSCAN_DB = metrics_kdb["DB"],
    K_DBSCAN_SI = metrics_kdb["SI"]
  ))
}

cat("\n========================================\n")
cat("FINAL METRICS COMPARISON (WITH NOISE):\n")
print(metrics_results)
```






```{r}
library(ggplot2)

# --- 1. Prepare Data for Davies-Bouldin Plot ---
df_db <- data.frame(
  N = rep(metrics_results$N, 2),
  Algorithm = rep(c("Standard DBSCAN", "K-DBSCAN"), each = nrow(metrics_results)),
  Score = c(metrics_results$DBSCAN_DB, metrics_results$K_DBSCAN_DB)
)

# Plot Davies-Bouldin
plot_db <- ggplot(df_db, aes(x = factor(N), y = Score, fill = Algorithm)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7, color="black") +
  scale_fill_manual(values = c("Standard DBSCAN" = "#F8766D", "K-DBSCAN" = "#00BFC4")) +
  labs(title = "Davies-Bouldin Quality Comparison",
       subtitle = "Lower values indicate better clustering separation",
       x = "Number of Data Points (N)",
       y = "Davies-Bouldin Index") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom",
        legend.title = element_blank())

# --- 2. Prepare Data for Silhouette Plot ---
df_si <- data.frame(
  N = rep(metrics_results$N, 2),
  Algorithm = rep(c("Standard DBSCAN", "K-DBSCAN"), each = nrow(metrics_results)),
  Score = c(metrics_results$DBSCAN_SI, metrics_results$K_DBSCAN_SI)
)

# Plot Silhouette
plot_si <- ggplot(df_si, aes(x = factor(N), y = Score, fill = Algorithm)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7, color="black") +
  scale_fill_manual(values = c("Standard DBSCAN" = "#F8766D", "K-DBSCAN" = "#00BFC4")) +
  labs(title = "Silhouette Quality Comparison",
       subtitle = "Values closer to 1 indicate better clustering cohesion",
       x = "Number of Data Points (N)",
       y = "Silhouette Index") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom",
        legend.title = element_blank())

# Print the plots to your viewer
print(plot_db)
print(plot_si)
```








